The JuliaHealth Observational Health Research Subecosystem

### 2024-01-01 paper.tex
Fixed up some more TODOs

% EXPLAIN::MEDIUM Why is this a crucial innovation?
% EXPLAIN::MEDIUM What is meant by a transferable analysis?
Arguably, the most crucial innovation that has emerged is the Observational Medical Outcomes Partnership Common Data Model (OMOP CDM) which was designed specifically to standardize observational health data to enable transferable analyses. \cite{overhage2012validation}
% EXPLAIN::MEDIUM Why is mapping relevant here?
The importance of this model has become apparent within open science organizations like the OHDSI (Observational Health Data Science and Informatics) network where members across $100+$ countries \cite{sachsonOurJourney2023} with access to various health systems have been able to successfully map their patient data to the OMOP CDM.
% TODO::EASY Add citation about successful collaborations
% EXPLAIN::EASY What is a network study?
% TODO::EASY Cite the quote "write once, run everywhere" 
% http://www.sun.com/smi/Press/sunflash/1996-01/sunflash.960123.10561.xml} 
This has led to successful collaboration relationships across countries where one research group can develop a network study and then disseminate the study across collaborators leading to the idea that you can "write once, run everywhere."

@@

% TODO::EASY Segue this better from phenotype definitions
One of the premier tools within the OHDSI HADES ecosystem is the ATLAS Shiny application which was developed to develop patient phenotype definitions within observational health research contexts (amongst other use cases).
These phenotype definitions have varying levels of complexity where they can be very simple (such as finding all patients with one health condition or diagnosis) or grow to be extremely complex (defining controlled, uncontrolled, and indeterminate patient cohorts based on medications, lab results, etc.)
% TODO::EASY Create footnote about schema
% TODO::EASY Add reference to schema in footnote
% TODO::EASY Cite reference schema from GitHub in footnote; \cite{madrilATLASJSONSchema2022}
% EXPLAIN::EASY Better introduce what SQL is
The computable phenotype definitions that ATLAS generates based on phenotype definitions are in a variety of serialization formats such as JSON or prepared SQL expressions available in multiple SQL flavors.

@@

% TODO::EASY Reference code repository
Using the JuliaHealth tools developed, we enumerate an example observational health study workflow that would be used in practice.\footnotemark

@@

% TODO::EASY Cite https://github.com/synthetichealth/synthea
\item \textbf{Eunomia} - this is a synthetic patient dataset created from the open source synthetic patient generator, \textit{Synthea}, and hosted within the HADES and JuliaHealth ecosystems.

@@

% TODO::EASY Mention in footnote explicitly that reproducing workflow in Eunomia is possible
% TODO::EASY Mention in footnote that explicitly reproducing workflow in MIMIC III is possible but not readily accessible
Reproducing the workflow using the \textit{Eunomia} dataset is made readily available at the repository here: \cite{schuemieEunomia2023}

% TODO::EASY Add link to code repository
\footnotetext{The code snippets presented in the body of the paper may omit certain intermediate steps. 
Please review the associated code repository for a full working tutorial located here: https://github.com/TheCedarPrince/2022-JuliaCon-Proceedings-Submission}

@@

\item Modularization of code and chaining

\item Distributed computing

\item Automatic generation of SQL from health packages

@@

% TODO::EASY Rename section
\subsubsection{Modularization of Code and Chaining}

@@

\subsubsection{Auto-Generating SQL from JuliaHealth Commands}

% TODO::EASY Rephrase this sentence
One crucial aspect of the OHS is that it is unavoidable to work with databases that contain patient health information.
% TODO::EASY Rephrase this sentence better
% TODO::EASY Cite ANSI SQL
For better or worse, the tool that is used to best interact with these databases are SQL dialects that loosely follow consistently the ANSI SQL guidance.
% EXPLAIN::MEDIUM Highlight this more
% EXPLAIN::MEDIUM In what scenarios might using Julia may not be possible
To accomodate for this, several of the core tools developed in this sub-ecosystem generate SQL that can be used in the case where Julia may not be available in a given environment housing patient health data.
% EXPLAIN::MEDIUM Introduce this better
As an example, \textit{OMOPCDMCohortCreator.jl} functions build upon \textit{FunSQL.jl} allowing one to use a function such `GetPatientAgeGroup` without a connection object which would result in SQL.

>>>>

% EXPLAIN::MEDIUM Why is this a crucial innovation?
% EXPLAIN::MEDIUM What is meant by a transferable analysis?
Arguably, the most crucial innovation that has emerged is the Observational Medical Outcomes Partnership Common Data Model (OMOP CDM) which was designed specifically to standardize observational health data to enable transferable analyses. \cite{overhage2012validation}
% EXPLAIN::MEDIUM Why is mapping relevant here?
The importance of this model has become apparent within open science organizations like the OHDSI (Observational Health Data Science and Informatics) network where members across $100+$ countries \cite{sachsonOurJourney2023} with access to various health systems have been able to successfully map their patient data to the OMOP CDM.
% TODO::EASY Add citation about successful collaborations
% EXPLAIN::EASY What is a network study?
This has led to successful collaboration relationships across countries where one research group can develop a network study and then disseminate the study across collaborators leading to the idea that you can "write once, run everywhere."\footnotemark

\footnotetext{This quote is attributed to Sun Microsystems when discussing the Java programming language.}

@@

% TODO::EASY Segue this better from phenotype definitions
One of the premier tools within the OHDSI HADES ecosystem is the ATLAS Shiny application which was developed to develop patient phenotype definitions within observational health research contexts (amongst other use cases).
These phenotype definitions have varying levels of complexity where they can be very simple (such as finding all patients with one health condition or diagnosis) or grow to be extremely complex (defining controlled, uncontrolled, and indeterminate patient cohorts based on medications, lab results, etc.)
% TODO::EASY Create footnote about schema
% TODO::EASY Cite reference schema from GitHub in footnote; 
% EXPLAIN::EASY Better introduce what SQL is
The computable phenotype definitions that ATLAS generates based on phenotype definitions are in a variety of serialization formats such as JSON or prepared SQL expressions available in multiple SQL flavors.\cite{madrilATLASJSONSchema2022}

@@

Using the JuliaHealth tools developed, we enumerate an example observational health study workflow that would be used in practice.\footnotemark

@@

\item \textbf{Eunomia} - this is a synthetic patient dataset created from the open source synthetic patient generator, \textit{Synthea}, and hosted within the HADES and JuliaHealth ecosystems. \cite{walonoski2018synthea}

@@

Reproducing the workflow using the \textit{Eunomia} dataset is made readily available at the repository here: \cite{schuemieEunomia2023}

\footnotetext{The code snippets presented in the body of the paper may omit certain intermediate steps. 
Please review the associated code repository for a full working tutorial located here: https://github.com/TheCedarPrince/2022-JuliaCon-Proceedings-Submission.
As a further note, the code to fully reproduce a sample analysis with \textit{Eunomia} is available within the code repository. 
As MIMIC III is not publicly available, a sample analysis using MIMIC III is available with some information stripped for privacy considerations.}

@@

\item Modularization

\item Distributed Computing

\item Automatic SQL Generation

@@

\subsubsection{Modularization}

@@

\subsubsection{Automatic SQL Generation}

% TODO::EASY Rephrase this sentence
One crucial aspect of the OHS is that it is unavoidable to work with databases that contain patient health information.
% TODO::EASY Rephrase this sentence better
For better or worse, the tool that is used to best interact with these databases are SQL dialects that loosely follow consistently the ANSI SQL guidance.\cite{kelechavaSQLStandardISO2018}
% EXPLAIN::MEDIUM Highlight this more
% EXPLAIN::MEDIUM In what scenarios might using Julia may not be possible
To accomodate for this, several of the core tools developed in this sub-ecosystem generate SQL that can be used in the case where Julia may not be available in a given environment housing patient health data.
% EXPLAIN::MEDIUM Introduce this better
As an example, \textit{OMOPCDMCohortCreator.jl} functions build upon \textit{FunSQL.jl} allowing one to use a function such `GetPatientAgeGroup` without a connection object which would result in SQL.
